Predicting Ambient Air Pollution (PM2.5) Across the Contiguous U.S.

### 2023-04-22 Draft.Rmd
Update Draft.Rmd

## Wrangling
```{r Predictors}
set.seed(123)
# Split origin dataset into 80/20
dat_split <- initial_split(origin, prop =.8)
dat_split

# Use 80% of data for training 
train <- training(dat_split)

# Create standardize train data for kNN and PCA
# excl. id and value
train_scaled = train[,3:50] |> dplyr::select(where(is.numeric)) |> mutate(across(fips:aod, scale)) |> cbind(state=train$state,county=train$county,city=train$city)

# Visualize linear relationship between predictors (lat, lon, state) and value
origin |> ggplot(aes(x=lon, y=value)) + 
  geom_point() + geom_smooth() + 
  labs(title="PM2.5 Concentration vs Longitude", x="Longitude (degrees)", y="PM2.5 (mu g/m^3)")

origin %>% ggplot(aes(x=lat, y=value)) + 
  geom_point() + geom_smooth() + 
  labs(title="PM2.5 Concentration vs Latitude", x="Latitude (degrees)", y="PM2.5 (mu g/m^3)")

origin %>% ggplot(aes(x=value, fill=state)) + geom_histogram() + facet_wrap(~state) + 
  labs(title="PM2.5 Concentration by State", x="PM2.5 (mu g/m^3)", y="Frequency (Monitors)") +
  theme(legend.position="none")
```

>>>>

## Wrangling
```{r Predictors}
set.seed(123)
# Split origin dataset into 80/20
dat_split <- initial_split(origin, prop =.8)
dat_split
# Use 80% of data for training 
train <- training(dat_split)

# Create standardize train data for kNN and PCA
# Omit value and id variables since these should not be scaled
train_scaled = train[,3:50] %>% dplyr::select(where(is.numeric)) %>% mutate(across(fips:aod, scale)) #%>% cbind(state=train$state,county=train$county,city=train$city)

# Calculate principal components on scaled training data
pca_result <- train_scaled %>% dplyr::select(where(is.numeric)) %>%
  prcomp()

# Plot scree plot
pca_result %>% fviz_eig()

# Plot table to show explained variance % by PCs
summary(pca_result)

# Threshold of Explained Variance = 80%
# Calculated the sum of PCs with highest explained variance until threshold was meant
0.3253 + 0.09749 + 0.07984 + 0.06501 + 0.05196 + 0.04147 + 0.03445 +
0.02926 + 0.02563 + 0.0220 + 0.02127 + 0.0188

PC1 <- pca_result$rotation[,1]
PC1_scores <- abs(PC1)
PC1_scores_ordered <- sort(PC1_scores, decreasing = TRUE)
names(PC1_scores_ordered)

# Reduce dimension of dataset
trunc <- pca_result$x[,1:12] %*% t(pca_result$rotation[,1:12])
# Add the center (and re-scale) back to data 
if(all(pca_result$scale != FALSE)){
  trunc <- scale(trunc, center = FALSE , scale=1/pca_result$scale) 
  } 
if(all(pca_result$center != FALSE)){ 
  trunc <- scale(trunc, center = -1 * pca_result$center, scale=FALSE) 
  } 
dim(trunc); dim(train_scaled)

# unhelpful plot -- delete
# plot(pca_result$x[,1], pca_result$x[,2])

# Show which features have the most influence
as_tibble(pca_result$rotation, 
          rownames = "variable") %>% 
    ggplot(aes(variable, PC1)) +
    geom_point() +
    coord_flip()

as_tibble(pca_result$rotation, 
          rownames = "variable") %>% 
    ggplot(aes(variable, PC2)) +
    geom_point() +
    coord_flip()

# Plot clusters between PC1 and PC2 using a biplot
fviz_pca_var(pca_result)

# Reduce dimension of dataset
trunc <- pca_result$x[,1:12] %*% t(pca_result$rotation[,1:12])
# Add the center (and re-scale) back to data 
if(all(pca_result$scale != FALSE)){
  trunc <- scale(trunc, center = FALSE , scale=1/pca_result$scale) 
  } 
if(all(pca_result$center != FALSE)){ 
  trunc <- scale(trunc, center = -1 * pca_result$center, scale=FALSE) 
  } 
dim(trunc); dim(train_scaled)

# Visualize linear relationship between predictors (lat, lon, state) and value
origin %>% ggplot(aes(x=lon, y=value)) + 
  geom_point() + geom_smooth() + 
  labs(title="PM2.5 Concentration (mu g/m^3)", x="Longitude (degrees)", y="PM2.5")

origin %>% ggplot(aes(x=lat, y=value)) + 
  geom_point() + geom_smooth() + 
  labs(title="PM2.5 Concentration (mu g/m^3)", x="Latitude (degrees)", y="PM2.5")

origin %>% ggplot(aes(x=value, fill=state)) + geom_histogram() + facet_wrap(~state) + 
  labs(title="PM2.5 Concentration (mu g/m^3)", x="State", y="PM2.5") 
```
